### 2020-08-26 thesis/appendix_1.tex
same with monteregie

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_39_compare.png}
}
 \caption{Comparison of Monteregie at the beginning of the BAU scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:BAU_compare}
\end{figure}

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_42_compare.png}
}
 \caption{Comparison of Monteregie at the beginning of the Reforestation scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:Ref_compare}
\end{figure}

@@

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_45_compare.png}
}
 \caption{Comparison of Monteregie at the beginning of the Corridor protection scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:Corr_compare}
\end{figure}

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_48_compare.png}
}
 \caption{Comparison of Monteregie at the beginning of the Corridor protection + Reforestation scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:	CorrRef_compare}
\end{figure}

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_51_compare.png}
}
 \caption{Comparison of Monteregie at the beginning of the Corridor protection + Targeted Reforestation scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:	CorrRefT_compare}
\end{figure}

>>>>

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_39_compare.png}
}
 \caption{Comparison of Montérégie at the beginning of the BAU scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:BAU_compare}
\end{figure}

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_42_compare.png}
}
 \caption{Comparison of Montérégie at the beginning of the Reforestation scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:Ref_compare}
\end{figure}

@@

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_45_compare.png}
}
 \caption{Comparison of Montérégie at the beginning of the Corridor protection scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:Corr_compare}
\end{figure}

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_48_compare.png}
}
 \caption{Comparison of Montérégie at the beginning of the Corridor protection + Reforestation scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:	CorrRef_compare}
\end{figure}

\begin{figure}[h!]
\makebox[\textwidth]{
  \includegraphics[width=1.3\textwidth]{thesis/figures/scenario_51_compare.png}
}
 \caption{Comparison of Montérégie at the beginning of the Corridor protection + Targeted Reforestation scenario in 2010, and at the end in 2100 (under baseline climate scenario).}
 \label{fig:	CorrRefT_compare}
\end{figure}

### 2020-08-26 thesis/chap_1.tex
same with monteregie

% [back to what we want to do]
Climate change can interact with land use change in ways that impact biodiversity \citep{oliver_interactions_2014} It is therefore likely that connectivity models integrating these two drivers would improve connectivity conservation methods such as CPAN design. Yet, in a recent review, \cite{costanza_landscape_2019} have found that connectivity methodological frameworks rarely combine  land use change and climate change. Here, we build on recent connectivity modelling work using circuit theory, by pairing it with a hybrid land use and climate change model \citep{albert_applying_2017}. The model we develop uses a combination of statistical modelling and simulations based on a Markov Chain Monte Carlo (MCMC) method, combining Random forests (RF) models with a Cellular automaton (CA) model. We use this RF-CA model to project future land use changes and estimate the changes in potential functional connectivity for five different umbrella species, under different climate change scenarios. We also derive potential insights for the design of a hypothetical regional network of connected protected areas resilient to future landscape change. In addition, we provide an analysis of past land use change trends in our study region, the Monteregie region in southern Quebec.

@@

% Review results and contrast with other papers integrating LUC and CC models
Other studies have attempted to go beyond those issues and managed to integrate climate and land use change models to some degree. For example, \cite{albert_applying_2017}, on which this work is based, assessed whether certain habitat networks in the Monteregie would be more resilient to future change using several multi scale metrics of connectivity. \cite{leonard_landscape-scale_2017} used systematic conservation planning methods to identify core areas and linkages under an energy development land use change scenario in the northeastern US. In another study, \cite{dilts_multiscale_2016} used a range of graph theoretic metrics to forecast the changes to diverse aspects of habitat connectivity for the Mohave ground squirrel under multiple climate and land use change scenarios. Although our results, contrary to those studies, do not allow us to derive prescriptive network design for our study species, our results do offer valuable insights for the design of such networks in the future.

>>>>

% [back to what we want to do]
Climate change can interact with land use change in ways that impact biodiversity \citep{oliver_interactions_2014} It is therefore likely that connectivity models integrating these two drivers would improve connectivity conservation methods such as CPAN design. Yet, in a recent review, \cite{costanza_landscape_2019} have found that connectivity methodological frameworks rarely combine  land use change and climate change. Here, we build on recent connectivity modelling work using circuit theory, by pairing it with a hybrid land use and climate change model \citep{albert_applying_2017}. The model we develop uses a combination of statistical modelling and simulations based on a Markov Chain Monte Carlo (MCMC) method, combining Random forests (RF) models with a Cellular automaton (CA) model. We use this RF-CA model to project future land use changes and estimate the changes in potential functional connectivity for five different umbrella species, under different climate change scenarios. We also derive potential insights for the design of a hypothetical regional network of connected protected areas resilient to future landscape change. In addition, we provide an analysis of past land use change trends in our study region, the Montérégie region in southern Quebec.

@@

% Review results and contrast with other papers integrating LUC and CC models
Other studies have attempted to go beyond those issues and managed to integrate climate and land use change models to some degree. For example, \cite{albert_applying_2017}, on which this work is based, assessed whether certain habitat networks in the Montérégie would be more resilient to future change using several multi scale metrics of connectivity. \cite{leonard_landscape-scale_2017} used systematic conservation planning methods to identify core areas and linkages under an energy development land use change scenario in the northeastern US. In another study, \cite{dilts_multiscale_2016} used a range of graph theoretic metrics to forecast the changes to diverse aspects of habitat connectivity for the Mohave ground squirrel under multiple climate and land use change scenarios. Although our results, contrary to those studies, do not allow us to derive prescriptive network design for our study species, our results do offer valuable insights for the design of such networks in the future.

### 2020-08-26 thesis/chap_2.tex
same with monteregie

This research is an extension of the work intitiated by the project “Monteregie Connection”, whose goal was to engage stakeholders around a discussion of the ecosystem services provided by natural habitat in the region. In the context of this stakeholder-driven project, \cite{mitchell_monteregie_2015} conducted a non-spatially explicit scenario building exercise for a subset of the Monteregie region, the Vallée-de-Richelieu Municipalité Region Comté. The outcome was 4 scenarios, ranging from “Periurban Development” to “Green Development”, and each describing different land use change trends. The Monteregie connection project improved our understanding of the relationship between forest connectivity and ecosystem services. However, there was no effort to estimate the impact of each scenario on ecological connectivity. This is partially due to a lack of tools able to translate these scenarios into simulations. We contribute to the development of such tools by integrating the results of the workshop into the land use change model developed in chapter 1. The goal of this chapter is therefore twofold: gain a better understanding of conservation priorities in the region, and to provide a basis for an iterative scenario-building process. \\
%-----------------------------------------------

@@

During our community workshop, participants identified connectivity conservation priorities in Monteregie. Participants quickly built consensus on which key cores areas and corridors should be prioritized for protection or restoration. They mainly identified large forested regions and protected areas as key cores areas. As key corridors, participants were strongly drawn to linear strips of connecting “stepping-stone” patches. They were rarely drawn to redundant corridor design. When participants were asked what factors would make a link a priority, they emphasized the quality of habitats to be linked, as well as a diversity of constraining and facilitating factors.

Participants also workshopped constraints and assets to connectivity conservation in Monteregie. We divided the results into elements that had been spatialized (i.e. placed on a specific location of the map during the exercise), and non spatialized (i.e. not placed on a specific location ), and scored them for the level of consensus they generated. Constraints scored higher, on average, than assets, indicating a focus of participants on constraints. The elements in the spatialized set scored higher on average than in the non spatialized set, indicating that participants tended to focus on more tangible and local issues and assets.

@@

The participants at the sub-regional (figure \ref{fig:reg_AC}) and regional (figure \ref{fig:trans_AC}) tables identified redundant groups of forest patches. Because regional tables covered the whole of Monteregie, they often identified cores that contained the cores identified by the sub-regional table. \\

@@

Conducting participatory research in Monteregie led us to a better understanding of the obstacles and opportunities for connectivity conservation in the region. The predominance of the agricultural land cover and its implications for land use planning and the regional economy was revealed as the most important obstacle for the region’s conservation goals, but was also seen as an opportunity to develop more biodiversity-friendly farming practices. The expansion of urban infrastructure also ranked high in more forested regions of the landscape. A range of assets was identified in the region: the strong engagement of local actors in connectivity conservation was a key factor. Through a consensus mapping workshop, we also mapped consensus priority core areas and links, showing how participants think corridors would make the best use of linear strips of the remaining forest in the region. Specifically, this demonstrated that participants saw small patches as important for corridor design and perceived corridors as an important tool to conserve and restore the potential of fragmented patches.

@@

The results of our simple scenarios have implications for connectivity conservation in the region. The reforestation scenario suggests that reforesting as much land as is urbanized is what is necessary to maintain mean current flow. But this is only one simple measure of connectivity, and therefore those results suggest that more work is needed to explore which metrics would be most appropriate to look at to support connectivity conservation planning, and more importantly what land use practices (protection, reforestation, urban densification) would be required to support the management of the ecological properties measured by this metric. In the light of the results from our workshop, it is clear that reforesting the same amount that we are urbanizing is possibly socially and economically unacceptable, given the amount of agricultural loss it generates. Even such a simple scenario echoes the long debated question of land sparing versus land sharing, a dichotomy that is now largely questioned as it fails to address the complexity of both food systems and the diversity of farming practices \cite{tscharntke_global_2012, bennett_changing_2017}. In fact, our reforestation scenario can be seen as an agricultural intensification model, in which farmers are able to cultivate less land to allow for more space for forest. Our workshop results suggest that stakeholders are instead likely to propose compensations to farmers for the loss in productive surface or for changes in practices (as was mentioned at some tables). This is likely reinforced by the fact that some stakeholders are opposed to further intensification of agricultural practices (also cited as an obstacle by participants). New compensation schemes such as ALUS (Alternative Land Use Services) are being experimented in Monteregie and are looking promising in their capacity to federare a bottom up movement of conservation in agrosystems \cite{ouellet_community_2020}.

The simple approach to the sparing vs sharing question that is brought about by our scenario overlooks the question of scale. At the scale of Monteregie, it is unclear to which extent the reduction in agricultural land can be promoted by the regional government. Like all modern farming systems, Québec's agriculture is embedded in a global market which might make it likely to prioritize global trends over local interests. Therefore, we are overlooking important variables for the design of credible alternative scenarios for conservation in the region, variables that are more likely to be elucidated in a participatory setting like the one we used.

>>>>

This research is an extension of the work intitiated by the project “Montérégie Connection”, whose goal was to engage stakeholders around a discussion of the ecosystem services provided by natural habitat in the region. In the context of this stakeholder-driven project, \cite{mitchell_monteregie_2015} conducted a non-spatially explicit scenario building exercise for a subset of the Montérégie region, the Vallée-de-Richelieu Municipalité Region Comté. The outcome was 4 scenarios, ranging from “Periurban Development” to “Green Development”, and each describing different land use change trends. The Montérégie connection project improved our understanding of the relationship between forest connectivity and ecosystem services. However, there was no effort to estimate the impact of each scenario on ecological connectivity. This is partially due to a lack of tools able to translate these scenarios into simulations. We contribute to the development of such tools by integrating the results of the workshop into the land use change model developed in chapter 1. The goal of this chapter is therefore twofold: gain a better understanding of conservation priorities in the region, and to provide a basis for an iterative scenario-building process. \\
%-----------------------------------------------

@@

During our community workshop, participants identified connectivity conservation priorities in Montérégie. Participants quickly built consensus on which key cores areas and corridors should be prioritized for protection or restoration. They mainly identified large forested regions and protected areas as key cores areas. As key corridors, participants were strongly drawn to linear strips of connecting “stepping-stone” patches. They were rarely drawn to redundant corridor design. When participants were asked what factors would make a link a priority, they emphasized the quality of habitats to be linked, as well as a diversity of constraining and facilitating factors.

Participants also workshopped constraints and assets to connectivity conservation in Montérégie. We divided the results into elements that had been spatialized (i.e. placed on a specific location of the map during the exercise), and non spatialized (i.e. not placed on a specific location ), and scored them for the level of consensus they generated. Constraints scored higher, on average, than assets, indicating a focus of participants on constraints. The elements in the spatialized set scored higher on average than in the non spatialized set, indicating that participants tended to focus on more tangible and local issues and assets.

@@

The participants at the sub-regional (figure \ref{fig:reg_AC}) and regional (figure \ref{fig:trans_AC}) tables identified redundant groups of forest patches. Because regional tables covered the whole of Montérégie, they often identified cores that contained the cores identified by the sub-regional table. \\

@@

Conducting participatory research in Montérégie led us to a better understanding of the obstacles and opportunities for connectivity conservation in the region. The predominance of the agricultural land cover and its implications for land use planning and the regional economy was revealed as the most important obstacle for the region’s conservation goals, but was also seen as an opportunity to develop more biodiversity-friendly farming practices. The expansion of urban infrastructure also ranked high in more forested regions of the landscape. A range of assets was identified in the region: the strong engagement of local actors in connectivity conservation was a key factor. Through a consensus mapping workshop, we also mapped consensus priority core areas and links, showing how participants think corridors would make the best use of linear strips of the remaining forest in the region. Specifically, this demonstrated that participants saw small patches as important for corridor design and perceived corridors as an important tool to conserve and restore the potential of fragmented patches.

@@

The results of our simple scenarios have implications for connectivity conservation in the region. The reforestation scenario suggests that reforesting as much land as is urbanized is what is necessary to maintain mean current flow. But this is only one simple measure of connectivity, and therefore those results suggest that more work is needed to explore which metrics would be most appropriate to look at to support connectivity conservation planning, and more importantly what land use practices (protection, reforestation, urban densification) would be required to support the management of the ecological properties measured by this metric. In the light of the results from our workshop, it is clear that reforesting the same amount that we are urbanizing is possibly socially and economically unacceptable, given the amount of agricultural loss it generates. Even such a simple scenario echoes the long debated question of land sparing versus land sharing, a dichotomy that is now largely questioned as it fails to address the complexity of both food systems and the diversity of farming practices \cite{tscharntke_global_2012, bennett_changing_2017}. In fact, our reforestation scenario can be seen as an agricultural intensification model, in which farmers are able to cultivate less land to allow for more space for forest. Our workshop results suggest that stakeholders are instead likely to propose compensations to farmers for the loss in productive surface or for changes in practices (as was mentioned at some tables). This is likely reinforced by the fact that some stakeholders are opposed to further intensification of agricultural practices (also cited as an obstacle by participants). New compensation schemes such as ALUS (Alternative Land Use Services) are being experimented in Montérégie and are looking promising in their capacity to federare a bottom up movement of conservation in agrosystems \cite{ouellet_community_2020}.

The simple approach to the sparing vs sharing question that is brought about by our scenario overlooks the question of scale. At the scale of Montérégie, it is unclear to which extent the reduction in agricultural land can be promoted by the regional government. Like all modern farming systems, Québec's agriculture is embedded in a global market which might make it likely to prioritize global trends over local interests. Therefore, we are overlooking important variables for the design of credible alternative scenarios for conservation in the region, variables that are more likely to be elucidated in a participatory setting like the one we used.

### 2020-08-26 thesis/thesis.tex
same with monteregie

Ecological connectivity is the extent to which the landscape supports the movements of organisms \citep{gonzalez_spatial_2017}, and is paramount for the resilience of both populations and ecosystem services \citep{mitchell_monteregie_2015} in heterogeneous and fragmented landscapes. Land use changes such as urban sprawl can cause deforestation, fragmenting habitats, and slowly eroding ecological connectivity. Many urban landscapes are experiencing uncontrolled urban sprawl and have suffered losses in connectivity and ecosystem services in consequence. Examples include cities like Barcelona \citep{marulli_gis_2005}, New York City \citep{mcphearson_urban_2014}, and also Montréal \citep{dupras_urban_2015}.  The drivers behind those land use changes are complex and understanding them is an obstacle to conservation planning  \citep{worboys_connectivity_2010}, including connectivity conservation planning. \\
Connectivity conservation planning (CCP), refers to the enterprise that engages multiple actors such as academics, NGOs, governmental bodies at different scales and in a common goal to conserve the ecological connectivity of the landscape. CCP methods usually entails modelling connectivity of the landscape of interest and using a prioritization method to determine conservation priorities, and modify land use practices to conserve or improve connectivity. This study constitutes a contribution to an ongoing effort of CCP in Montérégie, in southern Québec. In the following, we introduce the study region and describe previous efforts of CCP. We then point out ways in which CCP methods can be improved and how this thesis contributes to doing so. \\ \vspace{2em}

@@

The eastern part of Montérégie is also under the jurisdiction of the Greater Montréal Area (GMA). The ecological connectivity of the GMA and its benefits as a provider of ecosystem services has recently been assessed in a report to the Québec ministry of the environment (Rayfield 2018, unpublished). This study focused on identifying regions of highest connectivity, and therefore of highest priority for the conservation of biodiversity and ecosystem services. Other work by Rayfield et al. (2019, unpublished) has extended the analysis of connectivity to the whole of the Saint Lawrence Lowlands. Beyond connectivity the project of Monteregie Connection, spearheaded by an interdisciplinary team of researchers from McGill University, studied the provision of ecosystem services in the region and produced important knowledge for conservation planning. \\

Deriving insight from these contributions, multiple stakeholders have become mobilized to set connectivity conservation goals for the region. A number of Non-Governmental NGOs have designed corridor networks over their jurisdiction, with varying degrees of inputs from previous scientific work. For instance, Nature Action Québec (NAQ) and the Centre de la Nature du Mont Saint Hilaire (CDN), but also the Centre d'interprétation du milieu écologique du Haut-Richelieu (CIME), and Appalachian Corridors (CA), which has been pioneering the approach in the region (although their jurisdiction is at the margin of Monteregie, in Estrie). These associations are now beginning to work together to design science-driven connectivity conservation plans for the whole region, and are in need of decision tools in that regard. They are concerned with the social acceptability of conservation, and with resilience of their plan to future land use and climate change. \\ \vspace{2em}

@@

But, as we quoted from \cite{williamson_spatially_2018} above, this view tends to assume that stakeholders want to pursue rational and optimal design in their approach to conservation, and is once again a single-faceted approach to the problem, focusing on the trade-off between food production and support biodiversity. An alternative way is through the lens of ecosystem services  \cite{bennett_changing_2017}. Ecosystem services refer to the diversity of benefits that ecosystems provide to, and the many ways in which they support, human systems. Specifically, the concept of “bright spots” has potential for guiding the discussion beyond dichotomies and single faceted approaches, by focusing on multifunctionality in agroecosystems. In a landscape, “bright spots” are areas exceeding expectations for goals, such as agro-ecological multifunctionality, biodiversity, or connectivity \citep{frei_bright_2018}. They show promise as a tool  to identify levers of change, and guide stakeholder discussion. Indeed, an approach that focuses on identifying bright spots for the provision of those ecosystem services can be valuable to an interactive process of stakeholder engagement. They offer starting points to work with to design scenarios. They are concrete examples to reason around what bundles of ecosystem services are desirable (elements of SW) and which types of governance is needed to work toward them (elements of IC). In our study region, this framework is already being tested: \cite{frei_bright_2018} identified bright, dark and average spots within Monteregie for both landscape multifunctionality and avian biodiversity. They showed important spatial properties of bright spots, for instance that they were “neither spatially congruent nor in conflict”. \\

\citeauthor{frei_bright_2018} also pointed out a range of challenges for the identification of bright spots, the development of multicriteria measures being one of them. Connectivity measures are already being developed with  multifunctionality in mind. For instance, \cite{rayfield_priorisation_2018}, assessed whether certain habitat networks in the Monteregie would be more resilient to future change using several multi-scale metrics of connectivity. Another example is \cite{fletcher_towards_2019} who developed an unifying model for connectivity that takes into account multiple types of motability. Connectivity is therefore a field that fits in the multicriteria framework of bright spots, as it itself moves in this direction.  Therefore bright spots show promise for being integrated in more complex connectivity conservation planning tools. Finally, bright spots are intuitive, being based on statistical deviation from a given distribution and are therefore always relative to the context in which they are implemented. \\

>>>>

Ecological connectivity is the extent to which the landscape supports the movements of organisms \citep{gonzalez_spatial_2017}, and is paramount for the resilience of both populations and ecosystem services \citep{mitchell_Montérégie_2015} in heterogeneous and fragmented landscapes. Land use changes such as urban sprawl can cause deforestation, fragmenting habitats, and slowly eroding ecological connectivity. Many urban landscapes are experiencing uncontrolled urban sprawl and have suffered losses in connectivity and ecosystem services in consequence. Examples include cities like Barcelona \citep{marulli_gis_2005}, New York City \citep{mcphearson_urban_2014}, and also Montréal \citep{dupras_urban_2015}.  The drivers behind those land use changes are complex and understanding them is an obstacle to conservation planning  \citep{worboys_connectivity_2010}, including connectivity conservation planning. \\
Connectivity conservation planning (CCP), refers to the enterprise that engages multiple actors such as academics, NGOs, governmental bodies at different scales and in a common goal to conserve the ecological connectivity of the landscape. CCP methods usually entails modelling connectivity of the landscape of interest and using a prioritization method to determine conservation priorities, and modify land use practices to conserve or improve connectivity. This study constitutes a contribution to an ongoing effort of CCP in Montérégie, in southern Québec. In the following, we introduce the study region and describe previous efforts of CCP. We then point out ways in which CCP methods can be improved and how this thesis contributes to doing so. \\ \vspace{2em}

@@

The eastern part of Montérégie is also under the jurisdiction of the Greater Montréal Area (GMA). The ecological connectivity of the GMA and its benefits as a provider of ecosystem services has recently been assessed in a report to the Québec ministry of the environment (Rayfield 2018, unpublished). This study focused on identifying regions of highest connectivity, and therefore of highest priority for the conservation of biodiversity and ecosystem services. Other work by Rayfield et al. (2019, unpublished) has extended the analysis of connectivity to the whole of the Saint Lawrence Lowlands. Beyond connectivity the project of Montérégie Connection, spearheaded by an interdisciplinary team of researchers from McGill University, studied the provision of ecosystem services in the region and produced important knowledge for conservation planning. \\

Deriving insight from these contributions, multiple stakeholders have become mobilized to set connectivity conservation goals for the region. A number of Non-Governmental NGOs have designed corridor networks over their jurisdiction, with varying degrees of inputs from previous scientific work. For instance, Nature Action Québec (NAQ) and the Centre de la Nature du Mont Saint Hilaire (CDN), but also the Centre d'interprétation du milieu écologique du Haut-Richelieu (CIME), and Appalachian Corridors (CA), which has been pioneering the approach in the region (although their jurisdiction is at the margin of Montérégie, in Estrie). These associations are now beginning to work together to design science-driven connectivity conservation plans for the whole region, and are in need of decision tools in that regard. They are concerned with the social acceptability of conservation, and with resilience of their plan to future land use and climate change. \\ \vspace{2em}

@@

But, as we quoted from \cite{williamson_spatially_2018} above, this view tends to assume that stakeholders want to pursue rational and optimal design in their approach to conservation, and is once again a single-faceted approach to the problem, focusing on the trade-off between food production and support biodiversity. An alternative way is through the lens of ecosystem services  \cite{bennett_changing_2017}. Ecosystem services refer to the diversity of benefits that ecosystems provide to, and the many ways in which they support, human systems. Specifically, the concept of “bright spots” has potential for guiding the discussion beyond dichotomies and single faceted approaches, by focusing on multifunctionality in agroecosystems. In a landscape, “bright spots” are areas exceeding expectations for goals, such as agro-ecological multifunctionality, biodiversity, or connectivity \citep{frei_bright_2018}. They show promise as a tool  to identify levers of change, and guide stakeholder discussion. Indeed, an approach that focuses on identifying bright spots for the provision of those ecosystem services can be valuable to an interactive process of stakeholder engagement. They offer starting points to work with to design scenarios. They are concrete examples to reason around what bundles of ecosystem services are desirable (elements of SW) and which types of governance is needed to work toward them (elements of IC). In our study region, this framework is already being tested: \cite{frei_bright_2018} identified bright, dark and average spots within Montérégie for both landscape multifunctionality and avian biodiversity. They showed important spatial properties of bright spots, for instance that they were “neither spatially congruent nor in conflict”. \\

\citeauthor{frei_bright_2018} also pointed out a range of challenges for the identification of bright spots, the development of multicriteria measures being one of them. Connectivity measures are already being developed with  multifunctionality in mind. For instance, \cite{rayfield_priorisation_2018}, assessed whether certain habitat networks in the Montérégie would be more resilient to future change using several multi-scale metrics of connectivity. Another example is \cite{fletcher_towards_2019} who developed an unifying model for connectivity that takes into account multiple types of motability. Connectivity is therefore a field that fits in the multicriteria framework of bright spots, as it itself moves in this direction.  Therefore bright spots show promise for being integrated in more complex connectivity conservation planning tools. Finally, bright spots are intuitive, being based on statistical deviation from a given distribution and are therefore always relative to the context in which they are implemented. \\
